Algorithm explanation

### 2013-08-26 article/horn.tex
Other chapters also separated

\documentclass[master,final,12pt]{iscs-thesis}

\etitle{Simple Solutions of Acyclic Horn Clauses for Program Verification}
\eauthor{Yuto Takei}
\esupervisor{Naoki Kobayashi}

\jtitle{Lipsum}
\jauthor{Lipsum}
\jsupervisor{Lipsum}

\supervisortitle{Professor}
\date{August 28, 2013}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{stmaryrd}
\usepackage{syntax}
\usepackage{url}

\newcommand{\sembrack}[1]{[\![#1]\!}
\newcommand{\edge}[2]{#1\rightarrow#2}
\newcommand{\edgel}[3]{#1\xrightarrow{#2}#3}

\begin{document}

\begin{eabstract}
Automated theorem proving plays a crucial role in modern program
verification techniques.  In particular, in the context of software
model checking, the interpolation problem and its generalization
called \emph{Horn clause solving} have attracted special attention.
The latter problem asks for a solution for a set of acyclic Horn
clauses containing unknown predicate variables.  The solution, i.e., a
substitution for predicate variables, then serves as a candidate of
program invariants.  However, existing methods are unsatisfactory in
that they tend to generate too specific a solution, which does not
correspond to program invariants.

To address the problem above, we propose a new algorithm for solving
Horn clauses consisting of linear arithmetics and unknown predicate
variables.  The new algorithm tries to find small predicates, based on
the hypothesis that program's invariants tend to be simple formulas.
To this end, our algorithm prepares a template of linear arithmetic
formulas of a fixed size as a candidate of each unknown predicate, so
that Horn clause solving is reduced to linear programming.  The
algorithm iteratively increases the size of the templates until a
solution is found.  It enables us to construct relatively small
solutions in a reasonable time compared to existing algorithms.  We
have implemented and incorporated our Horn clause solver into MoCHi, a
software model checker for functional programs.  We have evaluated the
effectiveness of our approach through experiments.
\end{eabstract}
\begin{jabstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit.  Morbi
vulputate ornare arcu et malesuada.  Proin sem urna, feugiat et velit
ac, elementum blandit orci.  Etiam ut neque ac felis laoreet venenatis
eget quis enim.  Ut aliquam dapibus sem egestas molestie.  Ut sit amet
porttitor orci, id fermentum dui.  In non dui at ante sagittis
malesuada at in neque.  Fusce euismod velit quis ante ullamcorper
qsodales.
\end{jabstract}
\maketitle

\begin{acknowledge}
Thank you.
\end{acknowledge}

\frontmatter
\tableofcontents

\mainmatter

\chapter{Introduction}

Automated theorem proving plays an important role in the modern
program verification.  Particularly in the context of verification
with model checking techniques, an automated theorem prover is used to
abstract an infinite set of program states into a finite one so that
the model checker can virtually explore all possible program execution
paths.

One of the most common methods of such abstraction is predicate
abstraction\cite{conf/cav/GrafS97} within CounterExample-Guided
Abstraction Refinement (CEGAR) framework
\cite{conf/cav/ClarkeGJLV00,conf/popl/BallR02,conf/popl/HenzingerJMS02}.
An automated theorem prover finds appropriate predicates for every
program location.  An execution states at those locations is then
expressed by a Boolean valuation of the given predicates.  If the
abstraction is too coarse and a model checker discovers an infeasible
error execution path in an abstract system, the prover computes an
additional predicate at those program locations to refine the
abstraction predicates.

In order to obtain an additional abstraction predicate at program
locations along the infeasible counterexample, one class of automated
theorem provers, called an interpolating theorem prover
\cite{journals/tcs/McMillan05,conf/vmcai/RybalchenkoS07},
computes a separating predicate called an interpolant between two sets
of constraints; the one consists of the constraints from the program
entry point to the location, and the other consists of the constraints
from the location to the program failure.  Those constraints include
variable assignments and assertions to be satisfied originated from
conditional branches.

The computed predicate by an interpolating theorem prover is satisfied
at the location under the execution in the concrete model along the
discovered counterexample, and any execution paths which satisfies the
predicate does not lead to the failure state.  After the refinement,
the infeasible path is ruled out and no longer found by the model
checker.

However, existing interpolating theorem provers may return too complex
a solution which is heavily affected by the constraints of the
specific path.  It is not favorable in terms of the program
verification because a complex predicate may not be sufficient to rule
out similar spurious paths in programs with loops and recursion calls.
This causes the model checker to discover spurious paths infinitely
which pass the same program location by loops or recursion calls at
different numbers of iteration, and the verification process does not
to terminate.

To avoid this, it is desired to discover an invariant formula for the
predicate abstraction.  Based on the hypothesis that program's
invariants tend to be simple formulas, we propose a new interpolating
algorithm which tries to minimize the number of conjunctions and
disjunctions in a solution.  The predicate obtained by this algorithm
may become similar to the invariant, and the number of trials in model
checking may be reduced.

Additionally, we extend our interpolating algorithm to solve a
symmetric interpolation problem, in which the algorithm computes
predicates at multiple program locations along a path at once.  This
enables the abstraction predicates at multiple locations to be updated
at the same computation.

Finally, we extend our algorithm to solve Horn clause solving
problems.  It allows us to obtain predicates which are highly likely
to be program invariants, by computing the same abstraction predicates
inside loops and recursion calls on different iterations, and by
unifying predicates at the same location over multiple paths.  The
problems are generated by encoding constraints over execution paths
with unknown predicates which describe functions' constraints between
parameters and return values.

The rest of the paper is structured as follows.
Chapter~\ref{chap:interpolation} proposes a new interpolating
algorithm to obtain simple solutions, and extends it to solve
symmetric interpolation problems.  Chapter~\ref{chap:horn} describes a
new Horn clause solving algorithm by extending the previous
interpolating algorithm.  Chapter~\ref{chap:experiment} describes
experiments and illustrates the result to confirm the effect of our
proposed algorithm.  Chapter~\ref{chap:future} gives a discussion of
our improvements and proposes further improvements to make.
Chapter~\ref{chap:related} mentions related work done by others.
Finally, Chapter~\ref{chap:conclusion} reviews the impact of our
research and brings the conclusion.

\chapter{Interpolation}
\label{chap:interpolation}

Among various program verification techniques, when software model
checking with predicate abstraction is adopted, the interpolation is
used to compute the abstraction predicate along a spurious
counterexample that a model checker discovered.  It is done by
computing a Craig interpolant between the strongest postcondition at
the location and the weakest precondition to the failure point at
every program location.

However, by the existing interpolating theorem prover, we may not be
able to obtain a simple solution.  Consider the example below.
\begin{align*}
\varphi_1 : (x = 0 \wedge y = 0) \vee (x = 1 \wedge y = 1) \\
\varphi_2 : (x \neq 0 \wedge y = 0) \vee (x \neq 1 \wedge y = 1)
\end{align*}
Although one may be able to obtain $x = y$ as the simplest
interpolation between $\varphi_1$ and $\varphi_2$,
CSIsat\cite{conf/cav/BeyerZM08} returns
\[ TODO \]
as a solution.

We propose an interpolation algorithm which returns as simple a
solution as possible in terms of the number of conjunctions and
disjunctions.  Our algorithm focuses on interpolating problems on
linear arithmetic and it works in a constructive manner.

Consider the following interpolation problem.

\begin{align*}
\frac
{A: (x \leq a) \wedge (a + 1 \leq y)}
{B: (y \leq b) \wedge (b + 1 \leq x)}
\end{align*}

Here, the linear arithmetic formulas A and B are inconsistent.  The
interpolant I for this problem is $x-y+1 \leq 0$, which is implied by
$A$ and inconsistent with $B$.  The vocabulary of $I$ is $\left\lbrace
x,y \right\rbrace$ and is over $A$'s vocabulary $\left\lbrace x,y,a
\right\rbrace$ and $B$'s $\left\lbrace x,y,b \right\rbrace$.

In computing the interpolant, we make a linear constraint of a
interpolant by applying Farkas's lemma.  First, we assing weight
parameters to every expression in conjunction form.

\begin{align*}
\begin{array}{r r r r r r l}
\lambda_1 : & -a & & +x & & & \leq 0 \\
\lambda_2 : & a & & & -y & +1 & \leq 0 \\
\hline
\lambda_3 : & & -b & & +y & & \leq 0 \\
\lambda_4 : & & b & -x & & +1 & \leq 0
\end{array}
\end{align*}

Because all expressions induce inconsistent as a whole, the sum of
weighted expressions should satisfy the following constraint.

\begin{align*}
\begin{array}{r l l}
- \lambda_1 + \lambda_2 & = 0 \quad & (a) \\
- \lambda_3 + \lambda_4 & = 0 & (b) \\
  \lambda_1 - \lambda_4 & = 0 & (x) \\
- \lambda_2 + \lambda_3 & = 0 & (y) \\
  \lambda_2 + \lambda_4 & > 0 & (\text{constant}) \\
\lambda_1, \ldots, \lambda_4 & \geq 0 & (\text{non-negative weight})
\end{array}
\end{align*}

With the linear constraint among $\lambda_i$ above, the interpolant is
represented as follows by the sum of weighted expressions from the
first formula groups.
\[ (- \lambda_1 + \lambda_2) a + \lambda_1 x - \lambda_2 y + \lambda_2 \leq 0 \]
One of the model of the linear constraint is $\lambda_1 = \lambda_2 =
\lambda_3 = \lambda_4 = 1$, and we obtain $x-y+1 \leq 0$ as a
solution.

% TODO: Better to mention multiple interpolants possible for one
% interpolating problem?

Our algorithm aims to compute a Craig interpolant over linear
arithmetic formulas.

\paragraph{Craig Interpolation}
Given two logical formulas $A$ and $B$ that are inconsistent each
other, we call a new preposition $I$ a \emph{Craig interpolant}
\cite{journals/jsyml/Craig57} between $A$ and $B$ such that $I$
is implied by $A$ and inconsistent with $B$.  $I$'s vocabulary must be
only free variables that appears in both $A$ and $B$.  If $A \wedge B$
is unsatisfiable, an interpolant $I$ always exists.

Our algorithm handles a linear expression $e$ which is in a form:
\[ a_1 x_1 + \cdots + a_n x_n + b \leq 0 \]
This expression contains coefficient parameters $a_1, \ldots, a_n$ and
a constant parameter $b$.  We treat a special expressions $\top$ as $0
\leq 0$ and $\bot$ as $1 \leq 0$.  The algorithm input and output is a
formula $\psi$ that consists of conjunctions and/or disjunctions of
linear expressions.
\[ \psi = e | e \wedge e | e \vee e | \top | \bot \]

In the simple interpolation, the algorithm receives two formulas in
$\psi$ and returns an interpolant in $\psi$.  For the later extension
to handle a symmetric interpolation problem, the algorithm receives
multiple formulas in $\psi$ and returns intermediate interpolants in
$\psi$.  For computing those interpolants over linear arithmetics, we
use Farkas's Lemma.

\paragraph{Farkas's Lemma on linear inequalities}
Let a linear inequality $e_i$ be represented as $a_i1 x_1 + \cdots +
a_im x_m <= a_i0$.  Assuming that $e_1,\cdots,e_n$ implies $e_0$,
there exists $\lambda_1,\cdots,\lambda_n$ that satisfy $a_0j
=\sum_(i=1)^n \lambda_i * a_ij (j=0, \ldots, m)$.

For the special case of Farkas's Lemma, we can state that:

Assuming that $e_1,\cdots,e_n$ implies $\bot$, there exists
$\lambda_1,\cdots,\lambda_n$ that satisfy $???? TODO $.

Our interpolating algorithm make use of the latter version of Farkas's
Lemma.

\section{Algorithm}

Our algorithm aims to construct a relatively small interpolant in a
reasonable computation time for any input.  It also aims to obtain a
common solution among multiple interpolating problems.  When the
algorithm is used as a sub-routine for solving Horn clause satisfying
problems that are made while discovering predicates for abstraction
during the program verification.  These advantages of our algorithm
make it possible to return a relatively small predicates for those
problems. To accomplish these aims, the algorithm preserves a set of
interpolants during the computation, and define and execute operations
over interpolant sets.

\paragraph{Interpolation between conjunctions}
We first present an interpolating algorithm between two conjunctive
sets of linear inequalities $A = \left\lbrace e_1,\ldots,e_m
\right\rbrace$ and $B = \left\lbrace e_{m+1},\ldots,e_{m+n}
\right\rbrace$.  We assume that conjunctions of $A$ and $B$ are
inconsistent.  For later convenience, we call this interpolating
problem $\left( A, B \right)$.  Then there exists an interpolating
linear expression $e_\star$ which satisfies
\begin{align*}
\left\lbrace e_1,\ldots,e_m \right\rbrace & \vdash e_\star \\
e_\star & \nvdash \left\lbrace e_{m+1},\ldots,e_{m+n} \right\rbrace
\end{align*}
For convenience, we assume that any expression $e_i$ have variables
$x_1, \ldots, x_a$.

According to the Farkas's lemma, there exists a set of assignmentqs to
$\lambda_i$, to conclude $\bot$ from a set of linear expressions $A
\cup B$, namely,
\begin{align*}
\exists \lambda_1, \ldots, \lambda_m. \\
a % TODO: coefficient, constant constraints
\end{align*}

Then, a linear combination of $A$'s formula with a weight $\lambda_1,
\ldots, \lambda_m$ becomes an interpolant $I$ between $A$ and $B$.
Note that any assignments to $\lambda_i$ that satisfy the expressions
above will let an expression $I$ to be an interpolant.  Considering
this, it is possible to obtain different interpolants for a
interpolating problem by computing different assignments to
$\lambda_i$.

This enables the algorithm to preserve a set of interpolants by a
linear expression template $a_1 x_1 + \cdots + a_n x_n + b \leq 0$
with a constraint to describe a possible space of coefficient
parameters $a_i$ and the constant parameter $b$.

In practical, we may use linear programming solvers to obtain a model
for $\lambda_i$.  By assigning concrete values to them, we can obtain
a concrete interpolant.

\paragraph{Computing a common interpolant}
Next we consider a method to compute a common interpolant across two
different interpolating problems.  That is, when given two
interpolating problems $\left(A_1, B_1 \right)$ and $\left(A_2, B_2
\right)$, the problem is about to compute the common interpolant I,
which is an interpolant for the first problem and the second one at
the same time.

This problem is used when solving a following Horn clause solving
problem, for instance.
\[
\implies
% TODO
\]

To accomplish this, we simply need to compute the intersection of
interpolant sets.  Let us call the interpolant space for the problems
$\left(A_1, B_1 \right)$ and $\left(A_2, B_2 \right)$,
$\mathcal{I}_1$, and $\mathcal{I}_2$.  In the previous method, a set
of interpolants was represented by a parameterized linear expression.
\begin{align*}
\mathcal{I}_1 = \left\lbrace \mathbf{a}_1 \mathbf{x} \leq b_1 \mid
C_1 (\mathbf{a}_1, b_1 ) \right\rbrace \\
\mathcal{I}_2 = \left\lbrace \mathbf{a}_2 \mathbf{x} \leq b_2 \mid
C_2 (\mathbf{a}_2, b_2 ) \right\rbrace
\end{align*}
To compute the intersection between them, it is sufficient to choose
the same assignments of parameters and let them satisfy the both
constraints $C_1$ and $C_2$.
\[ \mathcal{I} = \left\lbrace \mathbf{a} \mathbf{x} \leq b \mid
C_1(\mathbf{a}, b) \wedge C_2(\mathbf{a}, b) \right\rbrace \]

However, if such an assignment of $\mathbf{a}$ and $b$ does not exist,
no common solution exists across two problems because $\mathcal{I} =
\emptyset$.

By applying this approach, it is also possible to compute a common
interpolant among more than three problems.

\paragraph{Interpolation between logical formulas}
We now extend the algorithm so that it can handle general logical
formulas over linear arithmetic including disjunctions of linear
expressions.  Here the algorithm takes two logical formulas over
linear expressions as a problem input.  For convenience, we assume
that they are represented in disjunctive normal form (DNF).  The
solution for this problem is the interpolant that is a logical formula
of linear expressions.

First, consider the case of an interpolation problem between $A_1 \vee
A_2$ and $B$.  When $(A_1 \vee A_2) \wedge B$ is inconsistent, $A_1
\wedge B$ is necessarily inconsistent.  Let $\mathcal{I}_1$ denote the
set of interpolants for $(A_1, B)$.  In the same manner, let
$\mathcal{I}_2$ denote the set of interpolants for $(A_2, B)$.  We may
naturally obtain an interpolant set by the intersection $\mathcal{I} =
\mathcal{I}_1 \cup \mathcal{I}_2$ as a solution for $(A_1 \vee A_2,
B)$.  When the algorithm computes $\mathcal{I}$ by the previous
method, and it is $\emptyset$, we need to build a less simple solution
by
\[ \mathcal{I}' = \left\lbrace \left( I_1 \vee I_2 \right) \mid
I_1 \in \mathcal{I}_1 \wedge I_2 \in \mathcal{I}_2 \right\rbrace \]
Any formulas in the set $\mathcal{I}'$ are the interpolant for $(A_1
\vee A_2, B)$ because
\begin{align*}
\forall I \in \mathcal{I}. & A_1 \vdash I \wedge A_2 \vdash I \wedge I \nvdash B \wedge \\
& FV(I) = FV(A_1) \cup FV(A_2) \cup FV(B)
\end{align*}

Second, consider the case of an interpolation problem between $A$ and
$B_1 \wedge B_2$.
% TODO:

By combining these two ways of interpolant construction, we may be
able to build a small solution for more general cases like $(A_1,
\ldots, A_n)$ and $(B_1, \ldots, B_n)$.  In order to obtain a small
solution, we need to wisely choose the order of combining an
interpolant set.

\paragraph{Symmetric Interpolation}
TODO

@@

By naturally encoding the first Horn clause with Farkas's lemma, we
obtain a linear constraint
$ \lambda_1 = c_1 \wedge - \lambda_1 \geq c_2 \wedge \lambda_1 \geq 0 $.
Any assignment to $ c_1, c_2, \lambda_1 $ which satisfies this will
make $A(x)$ valid for the first clause.  In the same manner, we obtain
$ \lambda_2 = c_3 \wedge - 2 \lambda_2 \geq c_4 \wedge \lambda_2 \geq 0 $
from the second clause, and
$ c_1 = 3 \lambda_3 \wedge c_3 = 2 \lambda_3 \wedge c_2 + c_4 \geq -10 \lambda_3 \wedge \lambda_3 > 0 $
from the third.

@@

traverses the input graph in the topological order on
$V_\Rightarrow$, and concatenate the constraints from predecessors to
build the problem constraints.  Each process to build a constraint on
$\mathcal{HC}_v$ for all vertices $v \in V_\Rightarrow$ is equivalent
to generate a linear constraint in interpolation.

\chapter{Experiment}
\label{chap:experiment}

We incorporated our algorithm to the software model checker for
higher-order programs (MoCHi) \cite{conf/pldi/KobayashiSU11}.  It
is specifically designed to verify the correctness of higher-orde
programs written in a small subset of ML.  It infers dependent types
for input higher-order programs \cite{conf/ppdp/UnnoK09}.

Our implementation make use of Z3 \cite{conf/tacas/MouraB08} as
an SMT solver.

For simplicity of the implementation, we assume that the linear
inequalities in the integer space.  Because of this, our algorithm
lost the completeness.  By using Motzkin's transpose theorem
\cite{journals/networks/Rajan90}, it is possible to handle linear
inequalities in both $\mathbf{ax} + b \leq 0$ and $\mathbf{ax} + b <
0$.

\section{Interpolating theorem prover}

Our interpolating theorem prover is used as a sub-procedure to
solve Horn clauses.

\begin{itemize}
\item CSIsat \cite{conf/cav/BeyerZM08}
\item FOCI \cite{website/foci}
\item CLP-Prover \cite{website/clp}
\end{itemize}

\section{Horn clause solver}

\chapter{Related work}
\label{chap:related}
\input{related.tex}

\chapter{Future Work}
\label{chap:future}

\paragraph{Non-linear extension}
\cite{conf/cav/DaiXZ13}

\chapter{Conclusion}
\label{chap:conclusion}

\bibliographystyle{plain}
\bibliography{./biblio}

\end{document}

>>>>

@@

By naturally encoding the first Horn clause with Farkas's lemma, we
obtain a linear constraint
$\lambda_1 = c_1 \wedge - \lambda_1 \geq c_2 \wedge \lambda_1 \geq 0$.
Any assignment to $ c_1, c_2, \lambda_1 $ which satisfies this will
make $A(x)$ valid for the first clause.  In the same manner, we obtain
$\lambda_2 = c_3 \wedge - 2 \lambda_2 \geq c_4 \wedge \lambda_2 \geq 0$
from the second clause, and 
$c_1 = 3 \lambda_3 \wedge c_3 = 2 \lambda_3 \wedge c_2 + c_4 \geq -10 \lambda_3 \wedge \lambda_3 > 0$
from the third.

@@

% TODO: What's this paragraph?
traverses the input graph in the topological order on $V_\Rightarrow$,
and concatenate the constraints from predecessors to build the problem
constraints.  Each process to build a constraint on $\mathcal{HC}_v$
for all vertices $v \in V_\Rightarrow$ is equivalent to generate a
linear constraint in interpolation.
